---
title: "TCRENV 2016"
author: "ICES Training Course in the R ENVironment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Out and in the open

Hopefully all source material is found [here](https://github.com/fishvice/tcrenv2016). If not mail us.

### Time and location

__Time__: 20160229 10:00 - 20160304 15:00

__Location__:
```{r, fig.width = 9, fig.height = 3, message = FALSE, echo=FALSE}
library(leaflet)
content <- paste(sep = "<br/>",
  "<b><a href='http://www.ices.dk'>ICES</a></b>",
  "H. C. Andersens Boulevard 44-46",
  "DK 1553 Copenhagen V, Denmark"
)
leaflet() %>% 
  addTiles() %>% 
  setView(lng = 12.575754, lat = 55.671455, zoom = 15) %>%
   addPopups(12.575754, 55.671455, content,
             options = popupOptions(closeButton = FALSE))
```

### Presentation
___

#### Preambles

* Presentation by ICES secretariat are located on the sharepoint

#### [Introduction to R and RStudio](pre/r-r-studio.pdf)

* [Rscript](pre/r-r-studio.R)

#### [Getting started with ggplot2](pre/ggplot_gettingstarted.html)

* [The source document](pre/ggplot_gettingstarted.Rmd)

#### [Reproducable research](pre/r-markdown.pdf)

* [Markdown example](pre/r-markdown-example.Rmd)

#### [Getting data into R](pre/r-data.pdf)

* [R-script](pre/r-data.R)

#### [Getting started with dplyr](pre/r-dplyr.pdf)

* [Rscript](pre/r-dplyr.R)

#### [Applied data and graph grammar](pre/more_grammar01.html)

* [The source document](pre/more_grammar01.Rmd)

#### [Fitting models](pre/r-models.pdf)

* [R-scripts](pre/r-models.R)

### Ad-hoc stuff
___

#### [Getting shapfiles into R](adhoc/shapefiles.html)

* [The source document](adhoc/shapefiles.Rmd)

#### [Nice catch colours](adhoc/Multispecies_IGFS_DATRAS_maps.html)

#### [The crayola survey](adhoc/crayola_survey.html)

* [The source document](adhoc/crayola_survey.Rmd)

#### [FLR %>% dplyr %>% ggplot](adhoc/flr.html)

* [The source document](adhoc/flr.Rmd)

#### [Small date drill](adhoc/datras_date_drill.html)

* [The source document](adhoc/datras_date_drill.Rmd)

#### [Generating a package](adhoc/rpackage_generation.html)

* [The source document](adhoc/rpackage_generation.Rmd)

### Potential projects
___

#### [DATRAS drills](prj/datras.html)

#### [ICES landings statistics](prj/ices_landings.html)

#### [EUROSTAT](prj/eurostat.html)

#### [Good environmental status](prj/GES.html)

#### [Working with logbooks data](prj/logbooks.html)

#### [Viewing stock summary statistics](prj/ss.html)

#### Your own data

### Overview of some "native" data used
___

#### [Icelandic spring groundfish survey](data_smb.html)

#### [Minke whales in Icelandic waters](data_minke.html)


### Post-hoc stuff
___

Material here added after the course

#### [Dplyr and databases](posthoc/databases.html)

