---
title: "Drilling DATRAS"
output: 
  html_document: 
    fig_height: 3
    fig_width: 5
---

## Preamble
___

[DATRAS](http://ices.dk/marine-data/data-portals/Pages/DATRAS.aspx) is a key fisheries science database used the within the ICES community.

It can be accessed either:

* From [DATRAS download page]https://datras.ices.dk/Data_products/Download/Download_Data_public.aspx) as an exchange format.
* Directly from within R via scripts that utilize the [DATRAS webservice](https://datras.ices.dk/WebServices/Webservices.aspx)

Some minor background information:

* Unique station identifier based on the following fields: year, quarter, country, ship, gear, stno, haulno
* In the field `lngtcode` in the length distribution file (filetype "HL") the "." and "0" mean that the length in the field `lngtclass` are in units of millimeters.
* The field `subfactor` is the raising factor

## Suggested procedures
___

* Utilize the provided `get_datras` function (see R directory on sharepoint) to read in the datras station (record type HH) and length data (record type HL) for a particular survey, year and quarter..
* Explore the data at your leasure.
* Suggested exercise (what we covered [here](http://www.hafro.is/~einarhj/education/tcrenv2016/pre/more_grammar01.html) might help):
    - Create a graphical map of the survey station location
    - Calculate the station abundance and biomass of some species of your choice and generate a map displaying the statistics
    - Calculate the number of station in each statistical rectangle, derive statistic of abundance and biomass and plot the results by statistical squares.
    - Obtain data over more years and calculate trends in abundance and biomass.

## Potential outcomes
___

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(ggplot2)
library(maps)
library(mapdata)
library(dplyr)
library(tidyr)
neamap <- 
  map_data("worldHires") %>%  
  filter((long > -50 & long < 50 & lat > 30) | region == "Greenland")
source("../R/datras.R")
source("../R/datras_tidy.R")
species <-
  read.csv("../data-raw/AphiaID_List.csv", stringsAsFactors = FALSE) %>%
  tbl_df() %>% 
  select(valid_aphia = WoRMS_AphiaID_Valid, latin = ScientificName_WoRMS) %>% 
  mutate(valid_aphia = as.integer(valid_aphia))
```

```{r, eval = FALSE}
# note this chunk not run - reduce traffic on datras
Station <- get_datras("HH", survey = "NS-IBTS", year = 2014, quarter = 1) %>% 
  tbl_df()
Length <- get_datras("HL", survey = "NS-IBTS", year = 2014, quarter = 1) %>% 
  tbl_df()
save(Station, Length, file="../data/nsibts2014q1.rda")
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
# ... instead of the above
attach("../data/nsibts2014q1.rda")
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
st <-
  Station %>% 
  tidy_station()
le <- tidy_length(Length, st) %>% 
  left_join(species) %>% 
  select(id, latin, sex, length = lngtclass, n)

background <- 
  list(theme_bw(),
       geom_polygon(data=neamap, aes(long, lat, group=group), fill = "grey90"),
       scale_x_continuous(NULL,NULL),
       scale_y_continuous(NULL,NULL))

# station plot
ggplot(st) +
  background +
  geom_segment(aes(shootlong, shootlat, xend=haullong, yend=haullat), col = "red") +
  coord_quickmap(range(st$shootlong), range(st$shootlat)) +
  labs(title = "NS-IBTS 2014, quarter 1")

# Calculate catch per haul
d <-
  le %>% 
  filter(latin %in% "Gadus morhua") %>% 
  mutate(wt = n * 0.01 * length^3) %>% 
  group_by(id) %>% 
  summarise(n = sum(n),
            b = sum(wt))
d2 <-
  st %>% select(id, shootlong, shootlat, statrec) %>% 
  left_join(d) %>% 
  mutate(n = ifelse(is.na(n), 0, n),             # if station is zero
         b = ifelse(is.na(b), 0, b)/1000)        # in kilos
ggplot(d2) +
  background +
  geom_point(aes(shootlong, shootlat, size = n), col = "red", alpha = 0.2) +
  scale_size_area(max_size = 10, breaks = c(0, 10, 50, 100, 200)) +
  coord_quickmap(range(d2$shootlong),range(d2$shootlat)) +
  labs(size = "n/haul", title = "NIBS 2014 Q1: Gadus morhua")

ggplot(d2) +
  background +
  geom_point(aes(shootlong, shootlat, size = b), col = "red", alpha = 0.2) +
  scale_size_area(max_size = 10, breaks = c(0, 10, 50, 100, 200))+
  coord_quickmap(range(d2$shootlong),range(d2$shootlat)) +
  labs(size = "kg/haul", title = "NIBS 2014 Q1: Gadus morhua")

ggplot(d2) +
  background +
  geom_point(aes(shootlong, shootlat, size = b/n), col = "red", alpha = 0.2) +
  scale_size_area(max_size = 10, breaks = c(0, 1, 2, 4, 8)) +
  geom_text(aes(shootlong, shootlat, label = round(n,0)), size = 2) +
  coord_quickmap(range(d2$shootlong),range(d2$shootlat)) +
  labs(size = "mean wgt", title = "NIBS 2014 Q1: Gadus morhua")


# Mean abundance in statistical rectangels
x <- d2 %>% 
  group_by(statrec) %>% 
  summarise(count = n(),
            n = mean(n),
            b = mean(b)) %>% 
  mutate(lon = geo::ir2d(statrec)$lon,
         lat = geo::ir2d(statrec)$lat)

ggplot(x) +
  geom_tile(aes(lon, lat, fill = n)) +
  background +
  geom_text(aes(lon, lat, label = round(n,0)), size = 2) +
  geom_point(data = d2, aes(shootlong, shootlat), col = "blue", size = 0.05) +
  scale_fill_continuous(low = "yellow", high = "red") +
  coord_quickmap(range(x$lon), range(x$lat)) +
  labs(fill = "n/rect", title = "NIBS 2014 Q1: Gadus morhua") +
  theme(legend.position = c(0.87,0.23))


x <- 
le %>% 
  filter(latin %in% "Gadus morhua") %>% 
  mutate(wt = n * 0.01 * length^3) %>% 
  group_by(length) %>% 
  summarise(n = sum(n),
            b = sum(wt)/1000) %>% 
  gather(key = variable, value = value, -length)


x %>% 
  ggplot() +
  theme_bw() +
  geom_area(aes(length, y = value, ymax = value, ymin = 0), fill = "grey") +
  facet_grid(variable ~ .) +
  labs(x = "Length [cm]", y = NULL, title = "Gadus morhua: numbers and mass")
```

