---
title: "Drilling euostat"
output: html_document
---

## Preamble
___

[Eurostat](http://ec.europa.eu/eurostat) contains wealth of informations, including statistics on [Fisheries](http://ec.europa.eu/eurostat/statistics-explained/index.php/Fishery_statistics), such as fleet capacity and catch by different nations.

A specific R-package, [eurostat](http://ropengov.github.io/r/2015/05/01/eurostat-package-examples) has been developed which allows one to access the database directly from within R.

## Suggested procedure
___

* Import the fleet ("fish_fleet") and catch tables ("fish_ca_main") using the function `get_eurostat`.
    - For the fleet table, decode the column "vessize" by using the function `label_eurostat`.
    - For the catch table, decode the column "species" by using the function `label_eurostat`.
* Try to repeat some of the analysis presented in [Fisheries statistics](http://ec.europa.eu/eurostat/statistics-explained/index.php/Fishery_statistics).
* Explore the data further at your leasure. E.g.
    - Analyse the changes in the fleet capacity over time by different nations
    - Analyse the changes in catch over time by different nations
    - Create some spatial plots (by country) of some statistics of interest
    - Calculate some index of fleet capacity in relation to the catch
    - ...


## Potential outcomes
___



```{r, message = FALSE, echo = FALSE}
library(eurostat)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(stringr)
```

```{r, eval = FALSE, echo = FALSE}
# note this chunk not run - reduce traffic on eurostat
Fleet <- get_eurostat("fish_fleet")
Catch <- get_eurostat("fish_ca_main")
save(Fleet, Catch, file = "data/eurostat.rda")
```

```{r, echo = FALSE}
#  instead of the above
attach("../data/eurostat.rda")
attach("../data/eumap.rda") # Einar: Have to find how obtained
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
fleet <- 
  Fleet %>% 
  mutate(vessize = as.character(vessize),
         code = label_eurostat(vessize, "vessize"),
         unit = as.character(unit),
         geo = as.character(geo),
         year = year(time),
         values = values/1e3) %>% 
  tbl_df() %>% 
  filter(nchar(geo) == 2,
         #!is.na(values),
         !str_detect(vessize, "_PC"))
fleetcode <- 
  fleet %>% 
  select(vessize, code) %>% 
  distinct() 

catch <- 
  Catch %>% 
  mutate(species = as.character(species),
         Species = label_eurostat(species, "species"),
         fishreg = as.character(fishreg),
         unit = as.character(unit),
         geo = as.character(geo),
         year = year(time),
         values = values/1e3) %>% 
  tbl_df() %>% 
  filter(nchar(geo) == 2)
speciescode <- 
  catch %>% 
  select(species, Species) %>% 
  distinct() 

fleet %>% 
  filter(unit == "GT",
         year == 2012,
         !is.na(values)) %>% 
  ggplot(aes(reorder(geo, values), values)) +
  geom_point() +
  coord_flip() +
  labs(x = NULL, y = NULL, title = "2012: Total fleet tonnage [thousands]")

fleet %>% 
  filter(unit == "KW",
         year == 2012,
         !is.na(values)) %>% 
  ggplot(aes(reorder(geo, values), values)) +
  geom_point() +
  coord_flip() +
  labs(x = NULL, y = NULL, title = "2012: Total kilowatt")

fleet %>% 
  filter(unit == "NR",
         year == 2012,
         !is.na(values)) %>% 
  ggplot(aes(reorder(geo, values), values)) +
  geom_point() +
  coord_flip() +
  labs(x = NULL, y = NULL, title = "2012: Total number")

fleet %>% 
  filter(vessize %in% c("T00")) %>% 
  group_by(geo) %>% 
  mutate(values = values/values[year == 2012]) %>% 
  ggplot() +
  geom_line(aes(year, values, colour = geo)) +
  geom_smooth(aes(year, values)) +
  facet_wrap(~ unit, scale = "free_y") +
  labs(x = NULL, y = NULL, title = "Development of fleet capacity")


gt <- fleet %>% 
  filter(year == 2010,
         vessize %in% c("T00"),
         unit == "KW")
gt %>% 
  ggplot(aes(map_id = geo)) +
  geom_map(aes(fill = values), map = eumap) +
  expand_limits(x = eumap$long, y = eumap$lat) +
  scale_fill_continuous(low = "yellow", high = "red") +
  labs(x = NULL, y = NULL, fill = NULL, title = "Fleet capacity: Kilowatts [thousands]")

ca <- 
  catch %>% 
  filter(species == "F00",
         fishreg == 0, 
         year == 2010)
ca %>% 
  ggplot(aes(map_id = geo)) +
  geom_map(aes(fill = values), map = eumap) +
  expand_limits(x = eumap$long, y = eumap$lat) +
  scale_fill_continuous(low = "yellow", high = "red") +
  labs(x = NULL, y = NULL, fill = NULL, title = "Total catch [thousand tonnes]")

ca %>% select(geo, catch = values) %>% 
  left_join(gt %>% select(geo, kw = values)) %>% 
  mutate(cpkw = catch/kw) %>% 
  ggplot(aes(map_id = geo)) +
  geom_map(aes(fill = cpkw), map = eumap) +
  expand_limits(x = eumap$long, y = eumap$lat) +
  scale_fill_continuous(low = "yellow", high = "red") +
  labs(x = NULL, y = NULL, fill = NULL, title = "Catch per 1000 kilowatts")
```

