---
title: "ICES stock assessment summaries"
output: html_document
---

A post-course project :-)

## Preamble
___

One of the backbone of ICES has been annual stock assessment and advice for miriads of stock. An overview of recent advisory documents can be found on the [ICES webpage](http://www.ices.dk/community/advisory-process/Pages/Latest-advice.aspx). The annual assessments have been compiled into a single worksheet (courtesy Martin Pastoors) found in the data-raw directory.

For the purpose of this excercise the column of interest are:

* __FishStock__: Name of the stock in ICES lingo
* __Species__: Short name of the species
* __Ecoregion__: The ecoregion that the stock falls under
* __AssYear__: The year the assessment was conducted
* __Year__: The year
* __F__: Fishing mortality
* __SSB__: The spawning stock biomass
* __Recruitment__: The recruitment abundance

## Suggested procedure
___

* Read in the data from the [excel document](../data-raw/ICES Assessment Summary database 20160303.xlsx)
* Plot the mortality trend by different ecoregions
* Plot some measure of the mean and variability of the terminal mortality estimates (that would be fishing mortality in year 2014) by e.g. ecoregion and/or some selected species.
* Tor each stock, calculate the ration of the contemporary fishing mortality and the fishing mortality estimated in the last assessment. Plot the retrospective patterns for some some selected species.
* Calculate a summary measure for each stock of the mean bias and the cv's based on the above statistic. Plot the results.
* Explore further at your leasure.

## Potential outcome
___

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
d <- read_excel("../data-raw/ICES\ Assessment\ Summary\ database 20160303.xlsx",
                sheet = "DATA") %>% 
  mutate(Ecoregion = ifelse(Ecoregion == "barents sea","Barents sea",Ecoregion))

d %>% 
  filter(AssYear %in% 2015) %>% 
  ggplot(aes(Year, F)) +
  geom_point(alpha = 0.3) +
  geom_smooth() +
  geom_vline(xintercept = 2014) +
  facet_wrap(~ Ecoregion) +
  labs(x = NULL, y = NULL, title = "Development of mortality")
````

```{r message = FALSE, warning = FALSE, echo = FALSE, fig.height = 2.5}
x <- d %>% 
  filter(AssYear %in% 2015,
         Year %in% 2014,
         !is.na(F),
         !is.na(Ecoregion))
source("../R/boot_trials.R")
x %>% 
  group_by(Ecoregion) %>% 
  summarise(bl = bootlower(F),
            bm = bootmean(F),
            bu = bootupper(F),
            n = n()) %>% 
  ggplot(aes(reorder(Ecoregion, bm), bm)) +
  geom_point(col = "red") +
  geom_linerange(aes(ymin = bl, ymax = bu), col = "red") +
  geom_text(aes(y = 0, label = n)) +
  coord_flip() +
  expand_limits(y = 0) +
  labs(x = NULL, y = NULL, title = "Mortality 2014 - bootrap mean and ci")
```

```{r message = FALSE, warning = FALSE, echo = FALSE}
stock2015 <- unique(d$FishStock[d$AssYear == 2015 & !is.na(d$F)])
ass2015 <-
  d %>% 
  filter(FishStock %in% stock2015,
         AssYear == 2015) %>% 
  select(Year, FishStock,
         F2015 = F,
         B2015 = SSB,
         R2015 = Recruitment)
x <- 
  d %>% 
  filter(FishStock %in% stock2015) %>% 
  left_join(ass2015) %>% 
  mutate(Frel = F/F2015,
         Brel = SSB/B2015,
         Rrel = Recruitment/R2015) %>% 
  select(Year, AssYear, FishStock, Species, Frel, Brel, Rrel) %>% 
  gather(key=variable,value=value, Frel:Rrel) %>% 
  filter(!is.infinite(value)) %>% 
  filter(!is.na(value)) %>%
  filter(value < 5)  # assume error

x %>% 
  filter(variable == "Frel",
         Species == "cod") %>% 
  ggplot(aes(Year, value, group = AssYear)) +
  geom_line() +
  facet_wrap(~ FishStock) +
  labs(x = NULL, y = "Fcontemporary/F2015",
       title = "Cod: mortality retrospective patterns")

x %>% 
  filter(variable == "Frel",
         Species == "sol") %>% 
  ggplot(aes(Year, value, group = AssYear)) +
  geom_line() +
  facet_wrap(~ FishStock) +
  labs(x = NULL, y = "Fcontemporary/F2015",
       title = "Sole: mortality retrospective patterns")

x2 <-
  x %>% 
  filter(Year == AssYear - 1,
         variable == "Frel") %>% 
  group_by(FishStock) %>% 
  summarise(bias = mean(value),
            cv = sd(value))
x2 %>% 
  ggplot(aes(reorder(FishStock, bias), bias)) +
  geom_point() +
  geom_hline(yintercept = 1) +
  coord_flip() +
  labs(x = NULL, y = NULL, title = "Mean bias in mortality")
x2 %>% 
  ggplot(aes(reorder(FishStock, cv), cv)) +
  geom_point() +
  coord_flip() +
  labs(x = NULL, y = NULL, title = "Mean cv in mortality")

```