---
title: "Getting started with ggplot2"
output: html_document
---

# Preamble

### Needed libraries for this tutorial:
```{r, message = FALSE, warning = FALSE, results='hide'}
library(geo)
library(ggplot2)
library(maps)
library(mapdata)
```

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE)
source("../R/common.R")
```

### Getting example data into R
```{r}
d <- read.csv("http://www.hafro.is/~einarhj/data/minke.csv",
              stringsAsFactors = FALSE)
d2 <- read.csv("http://data.hafro.is/assmt/2015/cod/summary.csv",
               stringsAsFactors = FALSE)
```

* __Get the data into your R-environment__

__The minke data__:

```{r}
str(d)
```

__The assessment summary data__:

Taken from <a href="http://data.hafro.is/assmt/2015/cod/" target="_blank">hafro data site</a>

```{r}
str(d2)
```

### Links

* The ggplot2 site: http://docs.ggplot2.org/current
* The ggplot2 book in the making: https://github.com/hadley/ggplot2-book
    - A rendered version of the book: http://www.hafro.is/~einarhj/education/ggplot2
        - needs to be updates
* R graphics cookbook: http://www.cookbook-r.com/Graphs/


# ggplot: Key components

ggplot has __three__ key components: 

1.  __data__,

2. A set of __aesthetic mappings__ [aes] between variables in the data and 
   visual properties, and 

3. At least one layer which describes how to render each observation. Layers
   are usually created with a __geom__ function.

`r columns(1)`
```{r}
ggplot(d, aes(x = age, y = length)) + 
  geom_point()
```

Different syntax, equivalent outcome:

```{r, eval = FALSE}
ggplot(d, aes(x = age, y = length)) + geom_point()
ggplot(d, aes(age, length))         + geom_point()
ggplot()                            + geom_point(data = d, aes(age, length))
ggplot(data = d)                    + geom_point(aes(x = age, y = length))
ggplot(d)                           + geom_point(aes(age, length))
d %>% ggplot()                      + geom_point(aes(age, length))
```

Can be stored as an object for later use:

```{r}
p <- ggplot(d, aes(age, length)) + geom_point()
```

The class:
```{r}
class(p)
```
The structure (a bit of latin - not run here):
```{r, eval = FALSE}
str(p)
```

# aesthetic: colour
`r columns(2)`
```{r}
p <- ggplot(d)
p + geom_point(aes(age, length, colour = sex))
p + geom_point(aes(age, length, colour = area))
```

Manual control of colours:
```{r}
p <- ggplot(d)
p + geom_point(aes(age, length, colour = sex)) +
  scale_colour_manual(values = c("orange","brown"))
p + geom_point(aes(age, length, colour = area)) +
  scale_colour_manual(values = c("orange","brown"))
```

# aesthetic: shape
`r columns(1)`
```{r}
p + geom_point(aes(age, length, shape = sex))
```

## Exercise

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Create a code that results in these plots:
`r columns(2)`
```{r, echo = FALSE}
p + geom_point(aes(age, length, colour = maturity))
p + geom_point(aes(age, length, shape = maturity))
```

</div>
</div>


# aesthetic: size
`r columns(1)`
```{r}
p + geom_point(aes(age, length, size = stomach.volume))
```

To reveal overlays:
`r columns(2)`
```{r}
p + geom_point(aes(age, length, size = stomach.volume), alpha = 0.6)
p + geom_point(aes(age, length, size = stomach.volume), alpha = 0.3, col = "red")
```

# facetting

Display same graph by subsets of the data:
`r columns(1)`
```{r}
ggplot(d) + 
  geom_point(aes(age, length, colour = sex)) + 
  facet_wrap(~ area)
```


<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create a code that results in this plot:
`r columns(1)`
```{r, echo = FALSE}
ggplot(d) +
  geom_point(aes(age, length, colour = area)) + 
  facet_wrap(~ sex)
```

</div>
</div>


# layers

## layer: add smoother to a plot

`r columns(2)`
```{r}
p <- ggplot(d, aes(age, length))
p + geom_point() + geom_smooth()
p + geom_point() + geom_smooth(span = 0.1)
```

Specific models:
`r columns(1)`
```{r}
p + geom_point() + geom_smooth(method = "lm")
```

## layer: points as jitters

Add a little random noise to the data to avoid overplotting:
`r columns(2)`
```{r}
p <- ggplot(d, aes(sex, length))
p + geom_point()
p + geom_jitter()
```

## layer: summarise via boxplots or violins
`r columns(2)`
```{r}
p + geom_boxplot()
p + geom_violin()
```

## layer: overlay data and summary layers

`r columns(1)`
```{r}
p + geom_boxplot() + geom_jitter()
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create the following 3 layer plot:

```{r, echo = FALSE}
p + geom_violin() + geom_boxplot() + geom_jitter()
```


</div>
</div>


## layer: frequency polygons

`r columns(2)`
```{r}
p <- ggplot(d, aes(length))
p + geom_freqpoly()
p + geom_freqpoly(aes(colour = sex))
```

## layer: histograms

More details distribution than boxplots do, but more space:
`r columns(2)`
```{r}
p + geom_histogram()
p + geom_histogram(aes(fill = sex))
```

Better way for histograms:
`r columns(1)`
```{r}
p + geom_histogram(aes(fill = sex)) + facet_wrap(~ sex, ncol = 1)
```

Controlling binwidths:
```{r}
p + geom_histogram(binwidth = 50)
```

## layer: bar charts

Histograms for discrete data:
`r columns(1)`
```{r}
ggplot(d, aes(maturity)) + geom_bar()
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Modify the above code to generate the following:

```{r, echo = FALSE}
ggplot(d, aes(maturity)) + geom_bar() + facet_wrap(~ area, nrow = 2)
```
</div>
</div>




## layer: line
`r columns(2)`
```{r}
ggplot(d2, aes(Year, Rec)) + geom_line()
ggplot(d2, aes(Year, Rec)) + geom_line() + expand_limits(y = 0)
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Create code to generate the following plots:
`r columns(2)`
```{r, echo = FALSE}
ggplot(d2, aes(Year, B4plus)) + geom_line() + expand_limits(y = 0)
ggplot(d2, aes(Year, Landings)) + geom_bar(stat = "identity")
```

The latter plot includes feature that has not been dealt with in these slides.
</div>
</div>


## layer: path
`r columns(2)`
```{r}
ggplot(d2, aes(SSB, Fbar)) + geom_line()
ggplot(d2, aes(SSB, Fbar)) + geom_path()
```

Check out difference of:
```{r, eval = FALSE}
?geom_line
?geom_path
```

Of more use:
`r columns(2)`
```{r}
ggplot(d2, aes(SSB, Fbar)) + geom_path() + geom_point(aes(colour = Year), size = 3)
```


# Some controls

## labels

`r columns(2)`
```{r}
p <- ggplot(d, aes(age, length, colour = sex)) + geom_point()
p
p + labs(x = "Age [year]", y = "Length [cm]", 
         colour = "Sex", title = "My minke plot")
```

## breaks

Controls which values appear as tick marks
`r columns(3)`
```{r}
p <- ggplot(d, aes(age, length)) + geom_point() + labs(x = NULL, y = NULL)
p
p +
  scale_x_continuous(breaks = c(5, 10, 15, 20, 25, 30, 35, 40, 45))
p +
  scale_x_continuous(breaks = seq(5, 45, by = 5)) +
  scale_y_continuous(breaks = seq(500, 950, by = 50))
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">
Create code that mimics the following breaks:
`r columns(2)`
```{r, echo = FALSE}
ggplot(d2, aes(Year, B4plus)) + 
  geom_line() + 
  expand_limits(y = 0) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  scale_y_continuous(breaks = seq(0, 2500, by = 250))
ggplot(d2, aes(Year, Landings)) +
  geom_bar(stat = "identity") +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10)) +
  scale_y_continuous(breaks = seq(0, 500, by = 100))
```


</div>
</div>



## limits

`r columns(3)`
```{r}
p <- ggplot(d, aes(maturity, length))
p + geom_jitter()
p + geom_jitter() + ylim(600, 800)
p + geom_jitter() + ylim(NA, 800) # setting only one limit
```

For descrete variables:
`r columns(1)`
```{r}
p + geom_jitter() + ylim(600,800) + xlim("immature","mature")
```

### warning

But be careful when using with summary statistics, e.g.:
`r columns(2)`
```{r}
p + geom_boxplot()
p + geom_boxplot() + ylim(600, 800)
```

Remedy:
`r columns(1)`
```{r}
p + geom_boxplot() + coord_cartesian(ylim = c(600, 800))
```

# A little gis

```{r}
str(island)
```

FAQ: where does island come from
A: check `?island`


`r columns(2)`
```{r}
p <- ggplot(island, aes(lon, lat)) + labs(x = NULL, y = NULL)
p + geom_point()
p + geom_line()
```

```{r}
p + geom_path()
p + geom_path() + coord_map()
```

## On maps

* Maps as background for r-plot can come from myriad of sources
* In this course we will largely focus on objects available in the map- and mapdata-packages.
* To get the data into ggplot2 friendly form (data.frame) we use the `map_data` function.

`r columns(1)`
```{r}
m <- map_data("world")
str(m)
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap()
```

The reslution of "word" is not very high as can be seen if only want iceland:
```{r}
m <- map_data("world", region = "denmark")
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap()
```

If we want higher resolution the object "worldHires" is often sufficient:
```{r}
m <- map_data("worldHires")
m <- m[m$long > -10 & m$long < 20 & m$lat > 50 & m$lat < 60,]
ggplot(m) +
  geom_polygon(aes(long, lat, group = group)) +
  coord_quickmap()
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Play with different map areas

</div>
</div>
