---
title: "Applied grammar"
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE, 
                      fig.width = 7.5,
                      fig.height = 4.5,
                      fig.show = "hold",
                      collapse = TRUE,
                      comment = "#>")
options(dplyr.print_min = 4L, dplyr.print_max = 4L)
```


# Preamble

* Here we want to enhance our skills in the use of grammar
* To that end we use some survey data
    * Icelandic groundfish survey
* We will take a step-by-step approach
* And end up with some catch per tow and mean abundance trends

# The data

For those that do not already know, procedure in a groundfish surveys goes quite often like this:

* Measurements of the catch composition is often based on a subsample of the total catch
* All species are identified and the total numbers/weight caught is recorded
* Length measurement of a species is quite often based on a subsample, the size of the subsample being species dependent (for some species all fish are measured, for some none).
* Age samples (something we are not going to worry about here) are quite often based on a smaller subsample than those only length measured.

Given the above the data are often stored in separate tables. Here we have three tables.

## Station table
see: http://www.hafro.is/~einarhj/data/tcrenv2016/Station.csv

Contains the following fields:

  * __id__: Unique station id
  * __date1__: Haul start time
  * __date2__: Haul end time
  * __lon1__ and __lon2__: Longitude of haul start and end location
  * __lat1__ and __lat2__: Latitude of haul start end end location
  * __towlengt__: Tow length in nautical miles
  * __depth__: Mean bottom depths in meters
  * __temperature__: Bottom temperature
  * __strata__: The strata of the tow

## Subsampling table
see: http://www.hafro.is/~einarhj/data/tcrenv2016/Subsampling.csv

Contains information about the subsampling at each station:

  * __id__: Station id
  * __species__: Species code
  * __n.counted__: Number of fish that were counted
  * __n.measured__: Number of fish that were measured
  * __n.total__: Total number of fish caught

## Length table:
see: http://www.hafro.is/~einarhj/data/tcrenv2016/Length.csv

Contains information about fish measured:

  * __id__: Station id
  * __species__: Species code
  * __length__: Length class in centimeters
  * __n__: Number of fish measured
  * __sex__: Sex code, females (2) and males(1)
  * __maturity__: Maturity code
  
# The code

### Needed libraries

```{r, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(geo)
library(maps)
library(mapdata)
```

### Recall the dplyr verbs (functions):

* __filter__: keep rows matching criteria
* __select__: pick columns by name
* __arrange__: order the rows according to a variable
* __mutate__: add new variables
* __summarise__: reduce variables to values
* __group_by__: Group data into rows
* __left_join__: return all rows from x, and all columns from x and y
* __right_join__: return all rows from y, and all columns from x and y

## Read in the data

The data needed are stored in three tables:
```{r readdata}
Station <- 
  read.csv("http://www.hafro.is/~einarhj/data/tcrenv2016/Station.csv") %>%
  mutate(year = lubridate::year(date1)) %>% 
  select(id, year, towlength, lon = lon1, lat = lat1) %>% 
  tbl_df()
Subsampling <- 
  read.csv("http://www.hafro.is/~einarhj/data/tcrenv2016/Subsampling.csv") %>% 
  tbl_df()
Length <- 
  read.csv("http://www.hafro.is/~einarhj/data/tcrenv2016/Length.csv") %>% 
  select(id, species, length, n) %>% 
  tbl_df()
```

For the record we have the following table dimensions:
```{r overview}
data_frame(table = c("Station","Length","Subsampling"),
           rows = c(nrow(Station),nrow(Length),nrow(Subsampling)),
           columns = c(ncol(Station),ncol(Length),ncol(Subsampling)))
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">


Familiarise yourself with / get an overview of the data. E.g. what is the year range the survey covers, how many stations per year, how many species are in the database, etc.

</div>
</div>

## A. One year, one species

#### 1. Pick a year and a species
```{r pick}
Species <- 1               # pick a species
Year <- 2015               # pick a year
```
#### 2. Subset (filter) the station table
```{r st}
st <-
  Station %>% 
  filter(year %in% Year) %>% 
  select(id, towlength, lon, lat) 
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* How many stations in the year you picked?
* Plot the towlength distribution
* Plot the tow locations

</div>
</div>


```{r st2, echo = FALSE, eval = FALSE}
nrow(st)
ggplot(st,aes(towlength)) + geom_histogram()
ggplot(st,aes(lon, lat)) + geom_point() + coord_map()
```

Actually since this survey is currently ongoing you can follow its progression [here](http://www.hafro.is/skip/skip.html).

#### 3. Subset (filter) the station table

* The year is not stored in the `Length` dataframe.
* So limit the data by using the station id in the `st` dataframe

```{r le}
le <-
  Length %>% 
  filter(id %in% st$id,
         species == Species) %>% 
  select(id, length, n)
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* How many length measurements?
* Plot the length distribution of fish measured

</div>
</div>

```{r le2, echo = FALSE, eval = FALSE}
sum(le$n)
ggplot(le, aes(length, n)) + geom_bar(stat = "identity")
p <- le %>% 
  group_by(length) %>% 
  summarise(n = sum(n)) %>% 
  ggplot(aes(length, n)) +
  scale_x_continuous(breaks = seq(0,140,10)) +
  expand_limits(x = 0)
p + geom_line()
p + geom_polygon(fill = "grey")
```

#### 4. Get the subsampling information
```{r, nu}
ss <-
  Subsampling %>% 
  filter(id %in% st$id,
         species == Species) %>% 
  select(id, n.total, n.measured)
```

#### ... and calculate the raising factor
```{r nu2}
ss <-
  ss %>% 
  mutate(r = n.total/n.measured,
         r = ifelse(is.na(r), 1, r)) %>%  # just in case
  select(id, r)                           # only columns needed
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* In how many cases (stations) were not all caught fish measured?

</div>
</div>

```{r nu3, echo = FALSE, eval = FALSE}
table(ss$r > 1)
```

#### 5. Raise the numbers measured

* Here we join the subsampling table (ss) with the length table (le):
```{r raise}
le <- 
  le %>% 
  left_join(ss) %>% 
  mutate(n = n * r)
```
* Filter the data such that you "see" the first few records were length was raised
```{r raise2, echo = FALSE, eval = FALSE}
le %>% filter(r > 1)
```

#### ... and calculate the weight
```{r weight}
le <-
  le %>% 
  mutate(wt = (n * 0.01 * length^3)/1000) # kilos
```

* Does the unit weight (kg) calculation make rough sense?

```{r sense, echo = FALSE, eval = FALSE}
le
```

#### 6. Calculate the catch per tow

* Each id represents a tow, hence we can group by it and then summarise

```{r pertow}
d <-
  le %>% 
  group_by(id) %>%
  summarize(n = sum(n),
            wt = sum(wt))
```

#### 7. Join the calculation with the station table

* We do this because the station table is the only place were the tow postion is recorded

```{r stationjoin}
d <-
  st %>% 
  left_join(d) %>% 
  mutate(n = ifelse(is.na(n), 0, n)/1e3,
         wt = ifelse(is.na(wt), 0, wt)/1e3)
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

* Why the ifelse mutate?
* Compare the numbers of records of dataframe `d` and `st`.

</div>
</div>

#### 8. Plot the results

```{r stationjoin2, fig.width = 7.5, fig.height = 4.5}
ggplot(d, aes(lon, lat)) + 
  geom_point(aes(size = wt), alpha = 0.3, colour = "red") +
  scale_size_area(max_size = 20, breaks = c(0, .1, .5, 1, 3)) +
  labs(x = NULL, y = NULL, size = "tonnes/tow") +
  coord_quickmap()
```

Here we would like to get some background to the map. Lets use the objects in the `mappackage`

```{r}
iceland <- map_data("worldHires", region = "iceland")
ggplot(d, aes(lon, lat)) + 
  theme_bw() +
  geom_polygon(data=iceland, aes(long, lat, group=group), fill = "grey90") +
  geom_point(aes(size = wt), alpha = 0.3, colour = "red") +
  scale_size_area(max_size = 20, breaks = c(0, .1, .5, 1, 3)) +
  labs(x = NULL, y = NULL, size = "tonnes/tow") +
  coord_quickmap()
```

#### A condensed script

```{r condensed, eval = FALSE}
Species <- 1
Year <- 2015
st <-
  Station %>% 
  filter(year %in% Year) %>% 
  select(id, towlength, lon, lat) 
d <-
  Length %>% 
  filter(id %in% st$id, 
         species %in% Species) %>% 
  select(id, length, n) %>% 
  left_join(Subsampling %>%                            # Note: here we do the raising
              filter(species %in% Species) %>%         #  "inside" the left_join function
              mutate(r = n.total/n.measured,
                     r = ifelse(is.na(r), 1, r)) %>% 
              select(id, r)) %>% 
  mutate(n = n * r,
         wt = (n * 0.01 * length^3)/1000) %>%     # kilos
  group_by(id) %>%                                # summarise per tow
  summarize(n = sum(n),
            wt = sum(wt)) %>% 
  right_join(st) %>%                              # join with station table
  mutate(n = ifelse(is.na(n), 0, n)/1e3,          #  note the right_join
         wt = ifelse(is.na(wt), 0, wt)/1e3)
```

#### Turning the script into a function

We are likely going to use this script often, e.g. for different year or different species. So we turn the above into a function.

* Note that the only thing changed here is the first line and the last line
    * In the first line we put the Year and Species inside the function call
    * In the last line we state explicitly return(d)
```{r function1}
catch_per_tow <- function(Year, Species) {
  st <-
    Station %>% 
    filter(year %in% Year) %>% 
    select(id, towlength, lon, lat) 
  d <-
    Length %>% 
    filter(id %in% st$id,             # this ensures we only work with SMB length 
           species %in% Species) %>% 
    select(id, length, n) %>% 
    left_join(Subsampling %>%
                filter(species %in% Species) %>%
                mutate(r = n.total/n.measured,
                       r = ifelse(is.na(r), 1, r)) %>% 
                select(id, r), by =  c("id")) %>% 
    mutate(n = n * r,
           wt = (n * 0.01 * length^3)/1000) %>%
    group_by(id) %>%
    summarize(n = sum(n),
              wt = sum(wt)) %>% 
    right_join(st, by = c("id")) %>%
    mutate(n = ifelse(is.na(n), 0, n)/1e3,
           wt = ifelse(is.na(wt), 0, wt)/1e3)
  
  return(d)
}
```

<div class="panel panel-warning">
<div class="panel-heading">You do</div>
<div class="panel-body">

1. Copy the above function code and put it a separate R-script.
    * In RStudio: File -> New file ... -> R script
2. Save the script as `catch_per_tow.R` in __your project R directory__.
3. When/if you want to use the function you simply run first:

```{r, eval = FALSE}
source("R/catch_per_tow.R")
```

</div>
</div>

#### Using the function

Here we pick another year and another species:
```{r}
catch_per_tow(Year = 2000, Species = 3) %>% 
  ggplot(aes(lon, lat)) + 
  theme_bw() +
  geom_polygon(data=iceland, aes(long, lat, group=group), fill = "grey90") +
  geom_point(aes(size = wt), alpha = 0.3, colour = "red") +
  scale_size_area(max_size = 20, breaks = c(0, .1, .25, 0.5, 1)) +
  labs(x = NULL, y = NULL, size = "tonnes/tow") +
  coord_quickmap()
```


<div class="panel panel-warning">
<div class="panel-heading">Exercises</div>
<div class="panel-body">

* Modify the function so that the catch per tow is expressed in standard towlength (4 nautical miles).
    * You may want to provide this as a users option.

</div>
</div>

#### Alternative representation

In many cases the mean abundance in statistical rectancles is calculate and displayed. To achieve this we need to:

* Asign each tow to statistical rectangle based on the lat and lons
    * Here we use the `d2ir` function in the geo-package
* Calculate the mean for each rectangle
* Get the cental position of the rectangles
    * Here we use the `ir2d` function in the geo-package
    
```{r}
d <- catch_per_tow(Year = 2015, Species = 2) 
d2 <- 
  d %>% 
  mutate(statrec = d2ir(lat, lon)) %>% 
  group_by(statrec) %>% 
  summarise(count = n(),
            n = mean(n),
            wt = mean(wt)) %>% 
  mutate(lon = geo::ir2d(statrec)$lon,
         lat = geo::ir2d(statrec)$lat)
```

And now for the plot (add here also number of station per rectangle and station location):
```{r}
ggplot(d2) +
  theme_bw() +
  geom_tile(aes(lon, lat, fill = wt)) +
  geom_polygon(data=iceland, aes(long, lat, group=group), fill = "grey90") +
  geom_text(aes(lon, lat, label = round(count,0)), size = 2) +
  geom_point(data = d, aes(lon, lat), col = "blue", size = 0.05, alpha = 0.5) +
  scale_fill_continuous(low = "yellow", high = "red") +
  coord_quickmap(range(d$lon), range(d$lat)) +
  labs(x = NULL, y = NULL, fill = "t/tow")
```

## B. Many years, one species

<div class="panel panel-warning">
<div class="panel-heading">Exercises</div>
<div class="panel-body">

* Modify the function such that it can handle multiple years.
* Create a bubble plot of some selected years (e.g. every tenth year) for some species

</div>
</div>

#### 1. Simple mean and median

```{r, echo = FALSE}
source("../R/catch_per_tow.R")
```

```{r}
d <- catch_per_tow(Year = 1985:2015, Species = 1) %>% 
  group_by(year) %>% 
  summarise(Median = median(wt),
            Mean = mean(wt)) %>% 
  gather(key = Statistic, value = value, -year)
d %>% 
  ggplot(aes(year, value, colour = Statistic)) +
  geom_point() +
  geom_line()
```

#### ... are the trends the same?

* Here we can simply standardize each statistics to the mean value

```{r ex5}
d %>% 
  group_by(Statistic) %>% 
  mutate(std.value = value/mean(value)) %>% 
  ggplot(aes(year, std.value, colour = Statistic)) +
  geom_line()
```

#### 2. Bootstrap mean and 95% confidence interval

* Here we can use the built in `mean_cl_boot` function

```{r}
catch_per_tow(1985:2015, 1) %>% 
  ggplot(aes(year, wt)) +
  stat_summary(fun.data = "mean_cl_boot", colour = "red") +
  expand_limits(y = 0) +
  labs(x = NULL, y = NULL, title = "Cod: Bootstrap mean biomass and confidence interval") +
  scale_x_continuous(breaks = seq(1985,2015,5))
```

```{r, eval = FALSE, echo = FALSE}
To obtain a tablular value of the bootstrap statistics
source("../R/boot_trials.R")
catch_per_tow(1985:2015, 1) %>% 
  group_by(year) %>% 
  summarise(bl = bootlower(wt),
            bm = bootmean(wt),
            lu = bootupper(wt))
```